### 2024-06-15 basic-statistics.qmd
20240612 - statistical decision tree
20240614 - research designs

@@

A Statistical Decision Tree For Choosing an Appropriate Statistical Procedure. Obtained from: <https://commons.wikimedia.org/wiki/File:InferentialStatisticalDecisionMakingTrees.pdf>. The original source is: Corston, R., & Colman, A. M. (2000). *A crash course in SPSS for Windows*. Wiley-Blackwell.
:::

>>>>

## Research Designs {#sec-researchDesigns}

There are three broad types of research designs:

- experiment
- correlational/observational study
- case study

### Experiment {#sec-experiment}

In an *experiment*, there are one or more things that we manipulate to see how the manipulation influences the process of interest.
The variable that we manipulate is the *independent variable*.
By contast, the *dependent variable* is the variable that we evaluate to determine whether it was influenced by the manipulation (i.e., by the independent variable).

For instance, we may be interested to evaluate whether players perform better (e.g., run faster) if they drink a sports drink compared when they drink only water.
Our hypothesis might be that players will be expected to perform better when they drink a sports drink (compared to when they drink only water).
To this this research question and hypothesis, we might conduct an experiment by randomly assigning some players during practice to receive a sports drink and some players to receive only water.
In this case, our independent variable is whether the player receives a sports drink.
Our dependent variable might be their 40-yard dash time during practice.

### Correlational/Observational Study {#sec-correlationalStudy}

In a correlational (aka observational) study, we do not manipulate a variable to see how the manipulation influences another variable.
Instead, we examine how two variables, a predictor and an outcome variable, are associated.
The hypothesized cause is called the predictor variable.
The hypothesized effect is called the outcome variable.
In this way, the predictor variable is similar to the independent variable, and the outcome variable is similar to the dependent variable.
However, unlike the independent and dependent variables in an experiment, the predictor and outcome variables in a correlational study are not manipulated.

For instance, to use a correlational study to test the possibility that players who drink sports drinks perform better than players who drink only water, we could examine whether the players who drink sports drinks during a game score more fantasy points than players who drink only water during the game.
In this case, our predictor variable is whether the players drinks sports drinks during a game.
Our outcome variable is the number of fantasy points the player scored.

### Case Study {#sec-caseStudy}

In a case study, we assess a small sample of individuals (commonly only one person or a few people), often with rich qualitative information.
Themes may be coded from the qualitative information, which may help inform inferences about whether some process may have played a role in influencing the outcome of interest.

For instance, to use a case study to evalute the possibilty that players who drink sports drinks perform better than players who drink only water, we could conduct an in-depth interview with a player.
In the interview, we might ask the player how they performed in games with versus without a sports drink and have them discuss whether they believe the sports drink improved their performance (and if so, how).
Then, based on the player's responses, we might code the responses to extract themes and to make a qualitative judgement of whether or not the player likely performed better during games in which they had a sports drinl

### Other Features of the Research Design: Number of Timepoints

In addition to whether the research design is an [experiment](#sec-experiment), [correlational/observational study](#sec-correlationalStudy), or a [case study](#sec-caseStudy), a research design can also have one or multiple timepoints.
The differing number of timepoints allow studies to be characterized as one of the following:

- cross-sectional
- longitudinal

#### Cross-Sectional {#sec-crossSectional}

A study with one timepoint is a *cross-sectional study*.

#### Longitudinal Design {#sec-longitudinal}

A study with more than one timepoint is a *longitudinal study*.
When the **same measures** are assessed at each of multiple timepoints, we refer to this as a "repeated measures" design.

## Research Design Validity {#sec-researchDesignValidity}

@@

A Statistical Decision Tree For Choosing an Appropriate Statistical Procedure. Adapted from: <https://commons.wikimedia.org/wiki/File:InferentialStatisticalDecisionMakingTrees.pdf>. The original source is: Corston, R., & Colman, A. M. (2000). *A crash course in SPSS for Windows*. Wiley-Blackwell. Changes were made to the original, including the addition of several statistical tests.
:::
